---
title: "Project Task #5, Initial ESDA"
author: "Reed Fu"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
html_document:
theme: darkly
highlight: breezedark
---

```{r setup, include=FALSE}
## knit option
knitr::opts_chunk$set(echo = TRUE)
## load in libraries
library(tidyverse)
library(sf)
library(tmap)
library(kableExtra)

## run prep code
## (To make sure everything runs, the setting of evaluate chunks in directory needs to be project)
# source("./code/data-prep.R")
```

******
### Read in Data Layers

```{r readin-data, results="hide"}
## read in sat data
nc_sat_21 <- read_csv("../data/process/2021-nc-sat.csv")
## read in shapefile
nc_school_dist_sf <- st_read("../data/process/nc-school-district.shp")
```

******
### Spatialize

```{r spatialize, results="hide"}

# Spatial join
nc_sat_21_sp <- merge(nc_school_dist_sf, nc_sat_21, by.x = "NAME", by.y = "district", all.x = TRUE)

# Aggregate schools by school district
nc_sat_21_dist_sp <- aggregate(nc_sat_21_sp[, c(4, 5, 6, 7)], by = list(nc_sat_21_sp$NAME), sum, na.rm=TRUE)

# Calculate average sat score
nc_sat_21_dist_sp$avg_combined <- round((nc_sat_21_dist_sp$total_math + nc_sat_21_dist_sp$total_erw)
                                          /nc_sat_21_dist_sp$num_tested)


```

******
### Data Description


```{r data-stat-1, eval = TRUE,  echo=FALSE, results="hide"}
n_obs <- nrow(nc_sat_21_dist_sp)
n_obs_na <- sum(is.na(nc_sat_21_dist_sp$avg_combined))
min_cmb_sat <- min(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
max_cmb_sat <- max(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
mean_cmb_sat <- mean(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
sd_cmb_sat <- sd(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
```


```{r data-stat-2, warning = FALSE, message = FALSE}
# Create summary information holder
summary.table <- tibble(Measure = c("Observations",
                                    "NA Observations",
                                    "Minimum",
                                    "Maximum",
                                    "Mean",
                                    "Standard Deviation"),
                        `Combined SAT Score` = c(n_obs,
                                                 n_obs_na,
                                                 min_cmb_sat,
                                                 max_cmb_sat,
                                                 mean_cmb_sat,
                                                 sd_cmb_sat),
)

## Print nice version of the table
kable(summary.table,
      digits = 0,
      format.args = list(big.mark = ",",
                         scientific = FALSE,
                         drop0trailing = TRUE),
      caption = "Summary of North Carolina Average Combined SAT Score by School District") %>%
  kable_styling(bootstrap_options = c("striped",
                                      "hover",
                                      "condensed",
                                      "responsive"),
                full_width = F)


```

There are total number of `r n_obs` number of observations,
out of those, `r n_obs_na` are NA observations.
The minimum average combined SAT score is `r min_cmb_sat`,
the maximum average combined SAT score is `r max_cmb_sat`,
the mean average combined SAT score is `r round(mean_cmb_sat)`, and
the standard deviation average combined SAT score is `r round(sd_cmb_sat)`.

Comments: Out of all school districts in NC,
there's quite a few (`r round(n_obs_na/n_obs * 100)`%) that don't have data.
For those districts that have data, the deviation is quite significant.

******
### Histogram

```{r histogram}
ggplot(nc_sat_21_dist_sp,
       aes(x = avg_combined)) +
  geom_histogram(binwidth = 15) +
  xlab("Average Combine SAT Scores in A School District") +
  ggtitle("Histogram of Public High School SAT Performance")
```

Comments: The histogram above shows the distribution of
average combined SAT performance by school district in North Carolina.
This histogram shows a symmetric and unimodal pattern as most school districts gather around 1130,
with minor very high and low scoring exceptions.


```{r Map-plotting}
# Create map
tm_shape(nc_sat_21_dist_sp) +
  tm_polygons("avg_combined",
              style = "jenks",
              palette = "Greens")
```

Comments:
The spatial pattern I see on the map is that there are three overall better performed regions.
Specifically in the east near Charlotte, in the middle around Triangle, and to the west near Wilmington.
This match my expectation, as more economically well-off school district tends to have more educational resources.
My hypothesis is that the there's a positive correlation with educational budget in a district with SAT performance.


******
### Summary

I learned that academic performance follows normal distribution very well when there's enough sample size.
There is a clear median and symmetrical distribution around it.
However, spatial representation in terms of choropleth map has a pattern
that there are higher level around metropolitan areas.
This suggests correlation between SAT performance to other social attributes
that could not be seen from data themselves.
The hypotheses I came up is that there the SAT performance might be related to other social factor,
for example income level or education budget.
It would be helpful to plot those factors to spatial maps and compare them side by side with SAT data.
