---
title: "task #3"
author: "Reed Fu"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    theme: darkly
    highlight: breezedark
---

```{r setup, include=FALSE}
## knit option
knitr::opts_chunk$set(echo = TRUE)
## load in libraries
library(tidyverse)
library(sf)
library(tmap)

## run prep code 
## (To make sure everything runs, the setting of evaluate chunks in directory needs to be project)
# source("./code/data-prep.R")
```

### Read in Data Layers

```{r readin-data}
  ## read in sat data
  nc_sat_21 <- read_csv("./data/process/2021-nc-sat.csv")
  ## read in shapefile
  nc_county_sf <- st_read("./data/process/nc-county.shp")
```

### Assess the Data Layers
```{r assess-data}
  ## glimpse sat data
  glimpse(nc_sat_21)
  ## glimpse shapefile
  glimpse(nc_county_sf)
```

The SAT data has `r nrow(nc_sat_21)` rows and `r ncol(nc_sat_21)` columns.
While the NC county shapefile has `r nrow(nc_county_sf)` rows 
and `r ncol(nc_county_sf)` columns.

### Plot a Map
```{r plot-map}
  plot(st_geometry(nc_county_sf))
```

### Minor Calculation
```{r sum-up}
  ## Add a new column of average total SAT based on average section scores
  nc_sat_21$avg_total_by_sum = nc_sat_21$avg_erw + nc_sat_21$avg_math
  ## the number of observations that the calculated total is different
  sum_diff <- nrow(filter(nc_sat_21, abs(avg_total_by_sum - avg_total) > 1))
```

The minor calculation I did is to create 
a new column of average SAT total scores (`$avg_total_by_sum`)
by adding up the average reading part (`$avg_erw`) 
and average math part (`$avg_math`).

This is useful when I calculate `sum_diff`, 
which is the number of observations 
that the supplied average total is varied from the calculated average total
for more than 1.
The reason I took the difference rather than directly compare is 
to minimize the effect of round off error.
And according to my code, there are `r sum_diff` observations 
show disparity among their data.
