---
title: "Spatial Distribution of SAT Score of North Carolina in 2021"
author:
  - Reed Fu  \newline
  - GEOG215, Introduction to Spatial Data Science
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    theme: darkly
    highlight: breezedark
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r setup, include=FALSE}
## knit option
knitr::opts_chunk$set(echo = TRUE)
## load in libraries
library(tidyverse)
library(sf)
library(sp)
library(spdep)
library(tmap)
library(kableExtra)
library(cowplot)

## (To make sure everything runs, the setting of evaluate chunks in directory needs to be project)
```

******
## Introduction

SAT is a widely used standardized test for college admission, 
and it's also a reflection of high school students' academic performance.
Analyzing the spatial pattern of SAT scores across the state 
provides some insight for understanding the public high school education in North Carolina.
The research question is: 
**Does the SAT performance of one school district in North Carolina correlates to its neighbors?**
To make sure all the code chunks and the data-prep R script run correctly, 
the knit directory has to be the _project directory_

******
## Data Preparation

### Sources
[SAT performance by school district in North Carolina in 2021](https://www.dpi.nc.gov/2021-sat-performance-district-and-school/open)
contains per school data for 
the total number of students, number of students who take the exam,
average score by exam section for all schools in the state of North Carolina in 2021.

The data is available as an Excel file provided by North Carolina Department of Public Instruction.
The Excel format was hard to deal with, specifically, the cascading layout of school district needs a work around.
First, I store school district as a separate column, then remove the column from the main table.
Then, I merge school district name back to the main table using district code.

I also remove all schools that do not belong to a school district. 
And convert all percentage data into absolute values so it would be easier to sum for later analysis.


[North Carolina school district shapefile](https://nces.ed.gov/programs/edge/data/EDGESCHOOLDISTRICT_TL21_SY2021.zip)
contains school district outlines in the entire country. 
It is downloaded from National Center for Education Statistics.
I have to subset the data by state code 37 to only keep school district in North Carolina,
and keep only the column of geoid, district name, and geometry. 

******
### Download, Process, and Read-in
The code chunk below downloads and prepare data for analysis.
```{r prepare-data, results="hide", message = FALSE, warning = FALSE}
## run data preparation code
# source("./code/data-prep-final.R")

## read in sat data
nc_sat_21 <- read_csv("./data/process/2021-nc-sat.csv", show_col_types = FALSE)

## read in shapefile
nc_school_dist_sf <- st_read("./data/process/nc-school-district.shp")
```

******
### Spatialization
This code chunk spatializes NC School SAT data with the school district shapefile.
```{r spatialize, results="hide", message = FALSE, warning = FALSE}
## Spatial join
nc_sat_21_sp <- merge(nc_school_dist_sf, nc_sat_21, by.x = "NAME", by.y = "district", all.x = TRUE)

## Aggregate schools by school district
nc_sat_21_dist_sp <- aggregate(nc_sat_21_sp[, c(4, 5, 6, 7)], by = list(nc_sat_21_sp$NAME), sum, na.rm=TRUE)

## Calculate average sat score
nc_sat_21_dist_sp$avg_combined <- round((nc_sat_21_dist_sp$total_math + nc_sat_21_dist_sp$total_erw)
                                          /nc_sat_21_dist_sp$num_tested)

## Calculate average erw score
nc_sat_21_dist_sp$avg_erw <- round(nc_sat_21_dist_sp$total_erw/nc_sat_21_dist_sp$num_tested)

## Calculate average math score
nc_sat_21_dist_sp$avg_math <- round(nc_sat_21_dist_sp$total_math/nc_sat_21_dist_sp$num_tested)
```

******
## Exploratory Spatial Data Analysis

### Description and Summary
```{r data-desc-sum-prep, eval = TRUE, results="hide", message = FALSE, warning = FALSE}
## Calculate summary information
n_obs <- nrow(nc_sat_21_dist_sp)
n_obs_na <- sum(is.na(nc_sat_21_dist_sp$avg_combined))
min_cmb_sat <- min(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
max_cmb_sat <- max(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
mean_cmb_sat <- mean(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
sd_cmb_sat <- sd(nc_sat_21_dist_sp$avg_combined, na.rm=TRUE)
```

```{r data-desc-sum-kable, warning = FALSE, message = FALSE}
## Create summary information holder
summary.table <- tibble(Measure = c("Observations",
                                    "NA Observations",
                                    "Minimum",
                                    "Maximum",
                                    "Mean",
                                    "Standard Deviation"),
                        `Combined SAT Score` = c(n_obs,
                                                 n_obs_na,
                                                 min_cmb_sat,
                                                 max_cmb_sat,
                                                 mean_cmb_sat,
                                                 sd_cmb_sat),
)

## Print nice version of the table
kable(summary.table,
      digits = 0,
      format.args = list(big.mark = ",",
                         scientific = FALSE,
                         drop0trailing = TRUE),
      caption = "Summary of North Carolina Average Combined SAT Score by School District") %>%
  kable_styling(bootstrap_options = c("striped",
                                      "hover",
                                      "condensed",
                                      "responsive"),
                full_width = F)
```

Out of all school districts in NC,
there's a few (`r round(n_obs_na/n_obs * 100)`%) that don't have data.
For those districts that have data, the deviation is quite significant.

*****
### Histogram
```{r data-desc-sum-hist, warning = FALSE, message = FALSE}
## Make a histogram of average combined score with bin-width 15
ggplot(nc_sat_21_dist_sp,
       aes(x = avg_combined)) +
  geom_histogram(binwidth = 15) +
  xlab("Average Combine SAT Scores in A School District") +
  ggtitle("Histogram of Public High School SAT Performance")
```

The histogram above shows the distribution of
average combined SAT performance by school district in North Carolina.
This histogram shows a symmetric and uni-modal pattern as most school districts gather around 1130,
with minor very high and low scoring exceptions.


******
## Geographic Distribution and Spatial Clustering
### Data Validation
```{r validation, results="hide", message = FALSE, warning = FALSE}
## Subset data to remove NA regions
nc_sat_21_dist_sp <- nc_sat_21_dist_sp[which(!is.na(nc_sat_21_dist_sp$avg_combined)), ]

## Check if all features are valid
valid_features <- sum(st_is_valid(nc_sat_21_dist_sp))

## Use an "if" statement to make valid, but ONLY IF NECESSARY
if (valid_features < nrow(nc_sat_21_dist_sp)) nc_sat_21_dist_sp <- st_make_valid(nc_sat_21_dist_sp)
```

******
### Data Mapping
```{r quick-map, message = FALSE, warning = FALSE}
## Tells tmap to create an interactive map
tmap_mode("view")

## Linked, side-by-side maps of SAT Reading-Writing and SAT math
tm_shape(nc_sat_21_dist_sp) +
  tm_basemap("CartoDB.PositronNoLabels") +
  tm_polygons(c("avg_erw", 
                "avg_math"), 
              style = "quantile", 
              border.col = "black", 
              border.alpha = 0.25) +
  tm_facets(sync = TRUE, 
            ncol = 2)
```

Both maps share similar patterns as there's a high scoring cluster to the south-west of the state near Charlotte,
and there's a low scoring cluster to the north-east.

******
### Moran's I
```{r morans-I, message = FALSE, warning = FALSE}
## Create Queen case neighbors
nc_sat_21_dist_sp_queen <- poly2nb(nc_sat_21_dist_sp, 
                       queen = TRUE)

## Convert the neighbors to weight matrix
nc_sat_21_dist_sp_weights <- nb2listw(nc_sat_21_dist_sp_queen, 
                          style = "B",         # B is binary (1,0)
                          zero.policy = TRUE)  # zero.policy allows for observations with NO neighbors 

## Moran's I
nc_sat_21_avg_combined_moran <- 
  moran.test(nc_sat_21_dist_sp$avg_combined,  # The column in your sp data
             nc_sat_21_dist_sp_weights,       # Weights object
             zero.policy = TRUE,              # Allows for observations with NO neighbors
             randomisation = TRUE)            # Compares to randomized NULL data

## Print summary to screen
nc_sat_21_avg_combined_moran
```

Moran's I analysis has a p-value of `r nc_sat_21_avg_combined_moran$p.value`,
which is very close to 0. 
The magnitude of Moran's I is `r nc_sat_21_avg_combined_moran$estimate['Moran I statistic']`,
which is also very close to 0.
Hence, the spatial correlation to it's neighbors are basically random.

******
### LISA
```{r LISA-calc, message = FALSE, warning = FALSE}
## LISA -- Local Moran's I
nc_sat_21_avg_combined_lisa <- 
  localmoran(nc_sat_21_dist_sp$avg_combined, # The column in your sp data 
             nc_sat_21_dist_sp_weights,      # Weights object
             zero.policy = TRUE) %>%         # Best to keep TRUE for LISA
  as.data.frame()  # Make result into data frame

## Scale the input data to deviation from mean
cDV <- nc_sat_21_dist_sp$avg_combined - mean(nc_sat_21_dist_sp$avg_combined)

## Get spatial lag values for each observation
## These are the neighbors' values!
lagDV <- lag.listw(nc_sat_21_dist_sp_weights, nc_sat_21_dist_sp$avg_combined)

## Scale the lag values to deviation from mean
clagDV <- lagDV - mean(lagDV, na.rm = TRUE)

## Add holder column with all 0s
nc_sat_21_avg_combined_lisa$Cat <- rep("0", nrow(nc_sat_21_avg_combined_lisa))


## This simply adds a label based on the values
nc_sat_21_avg_combined_lisa$Cat[which(cDV > 0 & clagDV > 0 & nc_sat_21_avg_combined_lisa[,5] < 0.05)] <- "HH" 
nc_sat_21_avg_combined_lisa$Cat[which(cDV < 0 & clagDV < 0 & nc_sat_21_avg_combined_lisa[,5] < 0.05)] <- "LL"      
nc_sat_21_avg_combined_lisa$Cat[which(cDV < 0 & clagDV > 0 & nc_sat_21_avg_combined_lisa[,5] < 0.05)] <- "LH"
nc_sat_21_avg_combined_lisa$Cat[which(cDV > 0 & clagDV < 0 & nc_sat_21_avg_combined_lisa[,5] < 0.05)] <- "HL"

```

```{r LISA-mapping, message = FALSE, warning = FALSE}
nc_sat_21_dist_sp$LISACAT <- nc_sat_21_avg_combined_lisa$Cat

## Tells tmap to create an interactive map
tmap_mode("view")

## Make a LISA map
tm_shape(nc_sat_21_dist_sp) + 
  tm_polygons("LISACAT", 
              style = "cat", 
              palette = c("grey",        # 0 (not significant)
                          "red",         # HH
                          "pink",        # HL
                          "lightblue",   # LH
                          "blue"),       # LL
              border.col = "Black", 
              border.alpha = 0.25) +
  tm_layout(legend.outside = TRUE) + 
  tm_basemap("CartoDB.PositronNoLabels")
```

Most school districts in North Carolina don't have spatial association that is significant enough.
The ones that show spatial association are entirely LL, which suggests there's a SAT performance swamp in the north-west of the state.
The cause of the cold spots could be further explored by comparing to other social factors.

******
## Conclusion

To sum up, there's hardly any spatial correlation of SAT performance of a school district in North Carolina to its neighbors. Both the numerical distribution and spatial distribution are close to random. However, there's several school districts that under perform compared to the rest of the state. More research could be done to investigate the social factors that might cause the cold spot.

******
#### Word Count
```{r word-count, echo = FALSE, message = FALSE}
wordcountaddin:::text_stats("./md/report.Rmd") %>% 
  kable_styling(bootstrap_options = c("striped", 
                                      "hover", 
                                      "condensed", 
                                      "responsive"), 
                full_width = F)
```
